### 2015-03-25 swmpr_manu.Rnw
finished complete first draft

%knitr options
<<setup, cache = F, echo = F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
options(replace.assign=TRUE,width=70,digits=1)
@

@@

The \acl{NERRS} (\acs{NERRS}, \url{http://www.nerrs.noaa.gov/}) is a network of 28 estuarine reserves in the United States that was established by the Coastal Zone Management Act of 1972.  The reserves represent different biogeographic regions and estuarine types that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.  As part of this effort, the \acf{SWMP} program was implemented in 1995 at over 300 stations at each of the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.  The \ac{SWMP} network has provided a continous source of data collected at near real-time at each of the reserves with the intent to evaluate natural and anthrogogenic causes of spatiotemporal variation in environmental condition and ecosystem function.  These data have been useful for evaluating relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).  However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \ac{SWMP} data.  \ac{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity of data provided by \ac{SWMP} \cite{SWMP14}.

This article describes a software package that was developed to address research needs of the \ac{NERRS} program using the open-source statistical programming language R \cite{RDCT14}.  SWMPr (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the System Wide Monitoring Program (\href{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}{SWMP}).  Functions provided by SWMPr address many of the common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify key parameters of interest.  Additionally, a cross-reserve comparison of current ecosystem metabolism estimates is provided to illustrate a potential application using the functions in this package.  The software is provided specifically for use with \ac{NERRS} data, although many of the applications are relevant addressing common challenges working with large datasets.

@@

\section*{Applications using the SWMPr package}

@@

The second and third examples are two Shiny web applications \cite{Chang15} that illustrate summaries and comparisons of \ac{SWMP} data within and between reserves that were created using the SWMPr package (see the \nameref{supp_info}).  The second application evaluates trends in \ac{SWMP} data within and between sites using an interactive map.  Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.  Site-levels trends are described below the map using a simple linear regression to show an increase or decrease in values over time.  Trends on the map at each station are plotted as circles that identify the direction and significance of the trend.  The trend direction is blue for decreasing and red for increasing.  The significance is indicated by radius of the circle and color shading where larger points with darker colors indicate a strong trend.  The third application provides graphical summaries of water quality, weather, or nutrient station data at individual stations. The drop down menus can be used to select the station, date range, and parameter for plotting. Plots are based on daily averages for each parameter.  The data used for each application are similar to those used to estimate ecosystem metabolism described above.

@@

Add some stuff....

@@

% Include only the SI item label in the subsection heading. Use the \nameref{label} command to cite SI items in the text.
\subsection*{Trends in \ac{SWMP} parameters}
\label{swmp_trends}
Evaluate trends within and between reserves using an interactive map: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}

>>>>

%knitr options
<<setup, cache = F, echo = F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F, background = 'white')
options(replace.assign=TRUE,width=70,digits=1)
@

@@

The \acl{NERRS} (\acs{NERRS}, \url{http://www.nerrs.noaa.gov/}) is a network of 28 estuarine reserves in the United States that was established by the Coastal Zone Management Act of 1972.  The reserves represent different biogeographic regions and estuarine types that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.  As part of this effort, the \acf{SWMP} program was implemented in 1995 at over 300 stations at each of the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.  The \ac{SWMP} network has provided a continous source of data collected at near real-time at each of the reserves with the intent to evaluate natural and anthrogogenic causes of spatiotemporal variation in environmental condition and ecosystem function.  These data have been useful for evaluating relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).  However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \ac{SWMP} data.  \ac{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by \ac{SWMP} \cite{SWMP14}.

This article describes a software package that was developed to address research needs of the \ac{NERRS} program using the open-source statistical programming language R \cite{RDCT14}.  SWMPr (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the System Wide Monitoring Program (\href{http://nerrs.noaa.gov/RCDefault.aspx?ID=18}{SWMP}).  Functions provided by SWMPr address many of the common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify key parameters of interest.  Additionally, a cross-reserve comparison of current ecosystem metabolism estimates is provided to illustrate a potential application using the functions in this package.  The software is provided specifically for use with \ac{NERRS} data, although many of the applications are relevant for addressing common challenges working with large datasets.

@@

\section*{Applications using the SWMPr package}
\label{swmp_apps}

@@

The second and third examples are two Shiny web applications \cite{Chang15} created using the SWMPr package that illustrate summaries and comparisons of \ac{SWMP} data within and between reserves (see the \nameref{supp_info}).  The second application evaluates trends in \ac{SWMP} data within and between sites using an interactive map.  Trends between reserves can be viewed using the map, whereas trends at individual sites can be viewed by clicking on a map location.  Site-levels trends are described below the map using a simple linear regression to show an increase or decrease in values over time.  Trends on the map at each station are plotted as circles that identify the direction and significance of the trend.  The trend direction is blue for decreasing and red for increasing.  The significance is indicated by radius of the circle and color shading where larger points with darker colors indicate a strong trend.  The third application provides graphical summaries of water quality, weather, or nutrient station data at individual stations. The drop down menus can be used to select the station, date range, and parameter for plotting. Plots are based on daily averages for each parameter.  The data used for each application are similar to those used to estimate ecosystem metabolism described above.

@@

The ability of management and research programs to address critical environmental issues is highly dependent on the quality of data used to inform decision making.  Standardized monitoring programs have vastly improved the ability to evaluate factors that influence a range of conditions, leading to more comprehensive assessment of site-specific characteristics and more informed decisions to manage environmental resources.  The \acl{SWMP} has provided twenty years of continuous monitoring of environmental characteristics at over over 300 stations within the 28 estuaries of the \acl{NERRS}.  This monitoring network establishes a foundation  for more effective coastal management by providing standardized data to address spatiotemporal variation in natural and anthropogenic characteristics that influence environmental condition.  Although the data provided by \ac{SWMP} are unique among coastal observing systems and have been used in a variety of applications \cite{Bulthius95,Caffrey03,Sanger02,Wenner04,Dix08}, the capacity of \ac{NERRS} researchers and staff to more effectively evaluate \ac{SWMP} data could be greatly improved using the SWMPr package to address common challenges working with large datasets.

The SWMPr package provides several functions to retrieve, organize, and analyze \ac{SWMP} data to more effectively evaluate environmental trends.  The package is designed to augment, rather than replace, existing data retrieval programs by providing a bridge betwen the raw data and the analysis software through its numerous data retrieval functions (\Cref{tab:retrieve}).  Established \ac{QAQC} methods and data processing techniques are also enhanced with \ac{SWMP} by functions that filter observations for different \ac{QAQC} flags (\texttt{qaqc}) and subset by selected dates or variables (\texttt{subset}).  Additionally, cumbersome challenges comparing differents datasets are addressed by the \texttt{setstep} and \texttt{comb} functions that standardize time steps and combine the data, respectively.  Finally, the analysis functions provide numerous tools to implement common analyses for time series and more specific methods for water quality data.  In particular, the \texttt{ecometab} function can be used to estimate daily integrated rates of ecosystem metabolism using the open-water method \cite{Odum56,Caffrey14}.  The above analysis (see \nameref{swmp_apps}) provided a cursory update of metabolism estimates for each the \ac{NERRS} estuaries using recent data to evaluate trends over time.  Although further evaluation of the data are needed, particularly regarding assumptions of the open-water method and tidal effects, the results could be used in a more comprehensive evaluation of factors that influence estuary metabolism. Further development of the SWMPr package will consider modifying existing and including additional functions to more effectively integrate data analysis with the quality of information provided by \ac{SWMP} and \ac{NERRS}.

@@

\subsection*{Trends in \ac{SWMP} parameters}
\label{swmp_trends}
Evaluate trends within and between reserves using an interactive map: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}
